Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 introduction.tex
made candidates more specific
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@874 ccdd05d5-56c1-461a-bf98-85f6876523b4
Changed some wording
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@875 ccdd05d5-56c1-461a-bf98-85f6876523b4

To this end, we have the ambition of reporting candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  By
looking for threshold crossings before the gravitational-wave signal leaves
the detection band, it is possible to trade some signal to noise ratio (\SNR{})
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for NS--NS binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources detectable within 10
seconds of merger and $\sim1$ sources detectable within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above \SNR{} 8.}
\end{figure}

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current trigger generation and telescope actuation would then dominate
latency.

Predictive detection of \CBC{}s will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm for
\textsc{fft} convolution, currently the preferred method in gravitational wave
searches.  The most obvious example of the latter is the time domain
(\textsc{td}) convolution, which is has no latency at all.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

>>>>

To this end, we have the ambition of reporting gravitational-wave candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  By
looking for threshold crossings before the gravitational-wave signal leaves
the detection band, it is possible to trade some signal to noise ratio (\SNR{})
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for NS--NS binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources detectable within 10
seconds of merger and $\sim1$ sources detectable within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above \SNR{} 8.}
\end{figure}

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current gravitational wave search methodology and telescope actuation would
dominate latency.

Advance detection of \CBC{}s will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm for
\textsc{fft} convolution, currently the preferred method in gravitational wave
searches.  The most obvious example of the latter is the time domain
(\textsc{td}) convolution, which is has no latency at all.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.
